A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.

### 2020-08-25 new-paper.Rmd
added footnotes (not needed anymore, but there it is)

---
title: A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.
author: Nicholas J Tierney\* (1,2) & Karthik Ram\* (3)
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: no
    includes:
      in_header: preamble.tex
bibliography: references.bib
header-includes: 
  - \setstretch{1.45}
  - \usepackage[utf8]{inputenc}
---

@@

Meaningful progress on computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by organizations like The Carpentries, combined with the growing adoption of Github by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely driven by the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python, R and Julia have helped scientists implement and share new methods to work with data [@rcore; @python; @julia]. Each of these languages enjoys support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interact with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, [@Collberg2014] and colleagues were only able to install and run less than 20% of the software described in publications. In addition to locating software, one still has to install the right versions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that trivial reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

@@

```{r incentable-table, fig.cap = "Table of the mechanism for behavior change, the incentive, and our assessment of where the elements of Data, Code, and Computational Environment rank in terms of completing these aspects. We note that Data is often required, but the preceding steps are not required. This is in contrast to Code having no policy"}
knitr::include_graphics("figures/gt-incentive-table.png")
```

@@

Tools to prepare datasets and metadata (*the infrastructure*) and best practices (guidelines) for reuse are sparse [@Arslan2019; @Roche2015]. Unlike code, datasets are far more diverse in terms of heterogeneity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. There are some efforts working towards improving this process, such as gogetdata , the 'dat' project, and a private company, Quilt [@goget; @dat; @quilt].

@@

Use a README file with your data project. A README is often the first place users will look to learn more about a directory [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how (<https://betta.io/blog/2017/02/07/developer-experience-github-readmes/> and [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards.

@@

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use a license from Creative Commons (<https://creativecommons.org/>), specifically, CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. Do not use non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

@@

Some data repositories and journals will automatically require a license when submitting data. For example, Dryad requires CC0 (<https://blog.datadryad.org/2011/10/05/why-does-dryad-use-cc0/> and <https://datadryad.org/stash/faq>) and the journal PLOS Comp Bio, requires licences for data that are not more restrictive than CC BY [@plos-comp-bio-data].

@@

**Variable names** are short, descriptive names with no spaces or special characters, such as "job\_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details, see [@Broman2017; @Ellis2017; @Arslan2019].

@@

Do not deposit data in local file storage that does not provide DOI or accession numbers. For example, [@50Hectares] was hosted on a university server and only accessible via a FTP server since 1982. It was not possible to easily access or cite the data without an explicit authorship agreement among other conditions. Only recently were the complete dataset deposited in Dryad without such conditions [@condit-2019].

@@

Journals provide data papers, where they provide a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad [@dryad] to link archived data with research papers. Dryad will also provide a person to help curate your data, improving usability.

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(rawdata)), or in software \@ref(data-as-part-of-open-source-packages). For medium data, consider a venue like zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.

>>>>

---
title: A Realistic Guide to Making Data Available Alongside Code to Improve Reproducibility.
author: Nicholas J Tierney\* (1,2) & Karthik Ram\* (3)
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: no
    includes:
      in_header: preamble.tex
bibliography: references.bib
csl: nature-communications.csl
header-includes: 
  - \setstretch{1.45}
  - \usepackage[utf8]{inputenc}
---

@@

Meaningful progress on computational reproducibility, the ability of an independent researcher to recreate figures, tables and key findings, became more of a practical reality in the late 2000s as new tools and services came into existence. Scientists started sharing more code and software, partly due to the rapid increase in training opportunities made possible by organizations like The Carpentries, combined with the growing adoption of Github by scientists [@Ram2013; @Zagalsky2015]. The bigger driver for this is more likely driven by the growing popularity of data science as a discipline distinct from statistics [@Donoho2017]. This rapid growth in data science has been catalyzed by a Cambrianesque explosion of open source software tools. Programming languages such as Python[^1], R and Julia have helped scientists implement and share new methods to work with data [@rcore; @julia]. Each of these languages enjoys support of thriving communities of researchers and software developers who contribute many of the building blocks that make them popular. Modern scientists now have access to many interactive computational frameworks (e.g. Jupyter, RMarkdown, Stencila, Observable) that can interact with kernels from dozens of programming languages. Beyond notebooks to run the code, a reader will also require access to the same software dependencies. In a study of applied computational research, [@Collberg2014] and colleagues were only able to install and run less than 20% of the software described in publications. In addition to locating software, one still has to install the right versions. Although this problem is far from solved, technologies like Docker have alleviated some of these concerns. Researchers can share Dockerfiles to recreate computational environments [@Boettiger2015]. Services like Binder can then leverage Docker and make possible to share the computational environments in which these notebooks were run. The barriers to verification are now low enough that trivial reproducibility is sometimes a click away, free from the misery of software installation [@Jupyter2018].

[^1]: <https://www.python.org/>

@@

```{r incentable-table, fig.cap = "Table of the mechanism for behavior change, the incentive, and our assessment of where the elements of Data, Code, and Computational Environment rank in terms of completing these aspects. We note that Data is often required, but the preceding steps are not required. This is in contrast to Code having no policy"}
library(gt)
incentive_table <- tibble(
  Mechanism = c("Policy", "Incentives", "Communities", "UI/UX", "Infrastructure"),
  Incentive = c("Required", "Rewarding", "Normative", "Easy", "Possible"),
  Data = c(100, 0, 0, 0, 100),
  Code = c(0, rep(100, 4)),
  `Computational Environment` = c(0, 0, 0, 50, 100)
  )

incentive_table %>%
  gt() %>%
  data_color(
    columns = vars(Data, Code, `Computational Environment`),
    colors = scales::col_numeric(
      # Using a function from paletteer to generate a vector of colors
      # Note that you need to wrap paletteer_d outputs in as.character()
      palette = colorspace::sequential_hcl(n = 3, palette = "Greens2", rev = TRUE),
      # Domain is from scales::col_numeric
      # Domain = The possible values that can be mapped
      # We don't HAVE to set a range for this since
      # we're using it inside data_color()
      domain = NULL
    )
  ) %>%
  gtsave("figures/gt-incentive-table.png")

# knitr::include_graphics("figures/gt-incentive-table.png")
```

@@

Tools to prepare datasets and metadata (*the infrastructure*) and best practices (guidelines) for reuse are sparse [@Arslan2019; @Roche2015]. Unlike code, datasets are far more diverse in terms of heterogeneity, size, and formats. This makes them particularly challenging to standardize or easily "install" where the code is running. There are some efforts working towards improving this process, such as gogetdata[^2] , dat[^3], and Quilt[^4].

[^2]: <https://gogetdata.github.io/>

[^3]: <https://dat.foundation/>

[^4]: <https://quiltdata.com/>

@@

Use a README file with your data project. A README is often the first place users will look to learn more about a directory [@Prana2019]. The README is a concise text file containing guidance on how to read and interpret the directory - the who, what, when, where, and how[^5] [@Prana2019]. For example, a README in a data repository might explain which directories contain data, tidying scripts, and also why data was collected. The README should be placed in the top level of the project, and there can be multiple README files in a project. In the context of datasets, READMEs are particularly useful when there are no reliable standards.

[^5]: <https://betta.io/blog/2017/02/07/developer-experience-github-readmes/>

@@

Provide a license with your data. A license for data clearly establishes the rules for its use, and whether it can be modified or shared. Without a license, or with a non-standard license, these rules are unclear, and lead to issues with use, attribution and citation. Use a license from Creative Commons[^6], specifically, CCBY and CC0 as they are commonly used, and provide simple rules for attribution, warranty, and use. Do not use non-standard licenses, such as those custom written by the authors, as some terms can render them unusable. Once a license is chosen, a LICENSE file should be provided, containing the entire license in the top level of the directory. The license type should be mentioned in the README, if used.

[^6]: <https://creativecommons.org/>

@@

Some data repositories and journals will automatically require a license when submitting data. For example, Dryad requires CC0[^7][^8], and the journal PLOS Comp Bio, requires licences for data that are not more restrictive than CC BY[^9].

[^7]: <https://blog.datadryad.org/2011/10/05/why-does-dryad-use-cc0/>

[^8]: <https://datadryad.org/stash/faq>

[^9]: <https://journals.plos.org/ploscompbiol/s/data-availability>

@@

**Variable names** are short, descriptive names with no spaces or special characters, such as "job_position" or "date". **Variable description** describes the measured variables, e.g., "University Job Position" or "Date Started", and should also define *variable codes* and *missing data*. **Unit text** defines the type of variable recorded (e.g., string, date, numeric, integer), and avoids serious errors of misrepresented data (e.g., a column of a gene sequence being interpreted as a date). For more details, see [@Broman2017; @Ellis2017; @Arslan2019].

@@

Do not deposit data in local file storage that does not provide DOI or accession numbers. For example, [@50Hectares] is hosted on a university server and accessed via FTP. It is not possible to effectively provide effective provenance and credit.

@@

Journals provide data papers, where they provide a formal publication explaining data that is archived. This is a familiar avenue for researchers to receive credit and citation. Writing a data paper can be very useful for researchers, however, data papers lack guidelines on how to structure data for effective reuse and sharing. Partner with Dryad[^10] to link archived data with research papers. Dryad will also provide a person to help curate your data, improving usability.

[^10]: <https://datadryad.org/>

@@

Small data has more options in how it is shared - you can bundle it with your analysis code (as in \@ref(rawdata)), or in software \@ref(pkgs). For medium data, consider a venue like zenodo for archiving. Sharing medium data might require it to be broken into separate chunks for reading (e.g., `teaching-1.csv`, `teaching-2.csv`, etc.), and scripts provided to read it. Large data will require liaising with services from \@ref(avenues), or local research institute library to make special arrangements, to ensure there is some way to gain access, and also maintain a DOI.
